Very simplified likelihood

### 2024-08-20 inst/hankin-AustrianJournalofStatistics.tex
more work on issue #239

Here a simplification of the Plackett--Luce likelihood functions for
order statistics is proposed.  We consider a field of $n+1$ entities
that are ordered in some way [the preferred example is competitors in
  a running race, ordered by time of crossing the finishing line].
One entity is of particular interest to us: the ``focal entity'', or
focal competitor.  We assign Plackett--Luce strength $a$ to the focal
competitor and $b$ to all others; we require $a+b=1$.  The associated
inference problem appears to have a wide range of applications and I
present some analysis of datasets drawn from the fields of Olympic
athletics, education, and Formula~1 motor racing.

@@

Ranked data arises whenever multiple entities are sorted by some
criterion; the preferred interpretation is a race in which the
sufficient statistic is the order of competitors crossing the
finishing line.  The Plackett--Luce likelihood
function~\citep{luce1959,plackett1975} is a generalization of the
Bradley--Terry model~\citep{bradley1952} in which non-negative
strengths are assigned to each competitor; without loss of generality,
if the finishing order is $1,2,\ldots,n$ and competitor $i$ has
strength $p_i\geqslant 0$, then the Plackett--Luce likelihood function
will be proportional to

@@

Consider a Plackett-Luce likelihood functions for a race between a
competitor of Bradley--Terry strength $a$ and $n\geqslant 1$ cloned
competitors, each of strength $b=1-a$.  Under these circumstances, an
observation is indexed by~$r$, the number of strength $b$ clones
finishing ahead of~$a$, and because we have $n+1$
competitors %%%%%%%%%%%%%%%%%%%%%% $0\leqslant r\leqslant n$.  The
initial field strength is $a+nb=n-(n-1)a$.  The Plackett--Luce
likelihood function~\citep{luce1959,plackett1975} for this situation
would be

@@

\noindent where $B=1/b$.  Note carefully that these expressions are
likelihoods, not probabilities, and do not sum to 1.  Asymptotic
analysis of expression~\ref{likeforrn3} shows that

\begin{equation}\label{asymptotic}
\log\mathcal{L}_{r,n}(a)=\log(B-1)-B\frac{r+1}{n}
+\mathcal{O}\left(n^{-2}\right)
\end{equation}

(up to an arbitrary additive constant).  Using
$\mathcal{S}'=\log(B-1)-B\frac{r+1}{n}$ as an approximate support thus
incurs an error of only $\mathcal{O}(n^{-2})$.  Observing that
$\partial^2\mathcal{S'}/\partial B^2=-(B-1)^{-2}<0$, the evaluate
would be unique; we have
$\hat{B}=\frac{n+r+1}{r+1}+\mathcal{O}\left(n^{-1}\right)$, or
alternatively $\hat{a}=\frac{n}{n+r+1}$.  Compare this with the result
of asking ``what is the probability that the focal competitor
beats a randomly chosen opponent?'' to which the binomial maximum
likelihood estimate is simply $\frac{n-r}{n}$, agreeing to within
$\mathcal{O}(n^{-2})$.  For repeated observations $r_i,n_i$,
$1\leqslant i\leqslant m$ we would have
\begin{equation}
  \hat{a} = 1-
  \frac{\frac{r_1+1}{n_1}+\cdots+\frac{r_m+1}{n_m}}{m}
  = 1-\overline{\left(\frac{r+1}{n}\right)}.
\end{equation}

@@

Previous related work \citep{hankin2024_hyper3} presented the case
$n=2$ [in current notation].  Considering the case $n=8$ gives us
$8+1=9$ distinct likelihood functions, Figure~\ref{ninelikes}.  We see
an unanticipated asymmetry between $r$ and $n-r$.  For $r=4$ one might
expect that $\hat{a}=\frac{1}{2}$ on the grounds that the focal
competitor's performance was median.  However, $\hat{a}\simeq 0.5886$.

@@

@@

One might wonder whether the student was better at pure or applied
mathematics and, if so, to quantify the difference.
Figure~\ref{plotpureandapplied} shows contours of support [again
  normalised so the support of the evaluate is zero] for the dataset,
as a function of pure strength on the horizontal axis and applied
strength on the vertical axis.  A null of equal strengths appears as
the diagonal line.  We see that the maximum likelihood estimate is
$\hat{p}_\mathrm{pure}=0.567$, $\hat{p}_\mathrm{applied}=0.887$.
However, a constrained optimization shows that the maximum support
along the null diagonal is about $-2.43$, at
${p}_\mathrm{pure}={p}_\mathrm{applied}=0.713$.  The two units of
support criterion~\citep{edwards1972} is met, and we are justified in
concluding that the student is indeed more competent at applied than
pure mathematics.  Alternatively, we may observe that
$2\mathcal{S}=4.86$ is in the tail region of its asymptotic $\chi^2_1$
distribution, corresponding to a $p$-value of about $0.028$.

>>>>

Here a simplification of the Plackett--Luce likelihood functions for
order statistics is proposed.  We consider a field of $n+1$ entities
that are ordered in some way [the preferred example is competitors in
  a running race, ordered by time of crossing the finishing line].
One entity is of particular interest to us: the ``focal entity'', or
focal competitor.  We assign Plackett--Luce strength $a$ to the focal
competitor and $b$ to all others; we require $a+b=1$.  The associated
inference problem has a wide range of applications and I present some
analysis of datasets drawn from the fields of Olympic athletics,
education, and Formula~1 motor racing.

@@

Ranked data arises whenever multiple entities are sorted by some
criterion; the preferred interpretation is a race in which the
sufficient statistic is the order of competitors crossing the
finishing line.  The Plackett--Luce likelihood
function~\citep{luce1959,plackett1975} is a generalization of the
Bradley--Terry model~\citep{bradley1952} in which non-negative
strengths are assigned to each competitor; without loss of generality,
if the finishing order is $1,2,\ldots,n$ and competitor $i$ has
strength $p_i\geqslant 0$, then the Plackett--Luce likelihood function
for this observation will be proportional to

@@

\subsection{Simplified Plackett-Luce likelihoods}

As noted, likelihood function~\ref{plackettluce} becomes unmanageable
with $n$ even moderately large and some form of simplification is
needed.  To that end, consider a Plackett-Luce likelihood function for
a race between a competitor of Bradley-Terry strength $a$ and
$n\geqslant 1$ cloned competitors, each of strength $b=1-a$.  Under
these circumstances, an observation is indexed by~$r$, the number of
strength $b$ clones finishing ahead of~$a$, and because we have $n+1$
competitors $0\leqslant r\leqslant n$.  The initial field strength is
$a+nb=n-(n-1)a$.  The Plackett--Luce likelihood
function~\citep{luce1959,plackett1975} for this situation would be

@@

\noindent where $B=1/b$.  Note carefully that these expressions are
likelihoods, not probabilities, and do not sum to 1.  Previous related
work \citep{hankin2024_hyper3} presented the case $n=2$ [in current
  notation].  Considering the case $n=8$ gives us $8+1=9$ distinct
likelihood functions, Figure~\ref{ninelikes}.  We see an unanticipated
asymmetry between $r$ and $n-r$.  For $r=4$ one might expect that
$\hat{a}=\frac{1}{2}$ on the grounds that the focal competitor's
performance was median.  However, $\hat{a}\simeq 0.5886$.
Figure~\ref{dotprobs} shows the maximum likelihood estimator $\hat{a}$
for various values of $n,r$, obtained by direct numerical maximization
of equation~\ref{likeforrn3}.

\subsection{Asymptotic analysis}

Analysis of expression~\ref{likeforrn3} for fixed $n$ shows that the
support $\mathcal{S}=\log\mathcal{L}_{r,n}(a)$ is asymptotically

\begin{equation}\label{asymptotic}
\mathcal{S}=
%\log\mathcal{L}_{r,n}(a)=
\log(B-1)-B\frac{r+1}{n}
+\mathcal{O}\left(n^{-2}\right)
\end{equation}

(up to an arbitrary additive constant).  Using
$\mathcal{S}^*=\log(B-1)-B\frac{r+1}{n}$ as an approximate
support thus incurs an error of only $\mathcal{O}(n^{-2})$.  Observing
that $\partial^2\mathcal{S}^*/\partial B^2=-(B-1)^{-2}<0$, the evaluate
would be unique; we have
$\hat{B}=\frac{n+r+1}{r+1}+\mathcal{O}\left(n^{-1}\right)$, or
alternatively $\hat{a}=\frac{n}{n+r+1}$.  Compare this with the result
of asking ``what is the probability that the focal competitor beats a
randomly chosen opponent?'' to which the binomial maximum likelihood
estimate is simply $\frac{n-r}{n}$, agreeing to within
$\mathcal{O}(n^{-2})$.  For repeated observations $r_i,n_i$,
$1\leqslant i\leqslant m$ we would have
\begin{equation}
  \hat{a} = 1-
  \frac{\frac{r_1+1}{n_1}+\cdots+\frac{r_m+1}{n_m}}{m}
  = 1-\overline{\left(\frac{r+1}{n}\right)}.
\end{equation}

@@

@@

\subsection{Park run}

``Parkrun'' is a distributed community initiative that organises
weekly timed 5 km runs/walks in parks worldwide~\citep{hindley2020}.
A typical event will have 200 participants.  Table~\ref{parkruntable}
shows the author has completed a total of 15 parkruns to date,
comprising 10 runs in Auckland and 5 at Stirling.  From the first pair
of numbers we see that 238 runners attended that particular parkrun,
of whom the author placed 173.

\begin{table}[t]
  \centering
  \caption{Parkrun results}
\label{parkruntable}
\begin{tabular}{ccccccccccc}
  \\
 & \multicolumn{10}{c}{Auckland}\\
rank    & 163  & 155 & 162 & 189 & 171 & 219 & 167 & 212 & 206 & 142 \\
runners & 238  & 238 & 196 & 242 & 242 & 318 & 259 & 305 & 297 & 241 \\ \\ \\
\end{tabular}
\begin{tabular}{cccccc}
  & \multicolumn{5}{c}{Stirling}\\
rank   & 118  & 224 & 128 & 115 & 183 \\
runners& 179  & 338 & 203 & 245 & 254 \\
\end{tabular}
\end{table}

Is there evidence that the author's performance differs between
Auckland and Stirling?  Figure~\ref{parkruncontour} shows contours of
support [again normalised so the support of the evaluate is zero] for
the parkrun dataset, as a function of Auckland strength on the
horizontal axis and Stirling strength on the vertical axis.  A null of
equal strengths appears as the diagonal line.  We see that the maximum
likelihood estimate is $\hat{p}_\mathrm{Auck}=0.427$,
$\hat{p}_\mathrm{Stir}=0.408$.  However, a constrained optimization
shows that the maximum support along the null diagonal is about
$-2.43$, at ${p}_\mathrm{Auck}={p}_\mathrm{Stir}\simeq 0.409$.  The
two units of support criterion~\citep{edwards1972} is met, and we are
justified in concluding that the student is indeed more competent at
applied than pure mathematics.  Alternatively, we may observe that
$2\mathcal{S}=4.86$ is in the tail region of its asymptotic $\chi^2_1$
distribution, corresponding to a $p$-value of about $0.028$.

\begin{figure}[t]
  \begin{centering}
\includegraphics[width=4in]{parkruncontour} % produced by very_simplified_likelihood.Rmd
\caption{Support contours for Bradley-Terry strength
  at \label{parkruncontour} Auckland and Stirling.  Null of equal
  strengths shown as a diagonal line}
\end{centering}
\end{figure}

@@

One might wonder whether the student was better at pure or applied
mathematics and, if so, to quantify the difference.
Figure~\ref{plotpureandapplied} shows contours of support [again
  normalised so the support of the evaluate is zero] for the dataset,
as a function of pure strength on the horizontal axis and applied
strength on the vertical axis.  A null of equal strengths appears as
the diagonal line.  We see that the maximum likelihood estimate is
$\hat{p}_\mathrm{pure}=0.567$, $\hat{p}_\mathrm{applied}=0.887$.
However, a constrained optimization shows that the maximum support
along the null diagonal is about $-2.43$, at
${p}_\mathrm{pure}={p}_\mathrm{applied}\simeq 0.713$.  The two units of
support criterion~\citep{edwards1972} is met, and we are justified in
concluding that the student is indeed more competent at applied than
pure mathematics.  Alternatively, we may observe that
$2\mathcal{S}=4.86$ is in the tail region of its asymptotic $\chi^2_1$
distribution, corresponding to a $p$-value of about $0.028$.

### 2024-08-20 inst/very_simplified_likelihood.Rmd
more work on issue #239

```{r,label=plotbothlikes}
par(pty='s')
contour(x,x,L,asp=1,xlim=range(x),ylim=range(x),levels=-(0:5))
abline(0,1)
```

@@

>>>>

```{r,label=plotbothlikes}
parkrun_contour <- function(...){
    par(pty='s')
    contour(x,x,L,asp=1,xlim=range(x),ylim=range(x),levels=-(0:5),xlab="Auckland",ylab="Stirling")
    abline(0,1)
}
parkrun_contour()
pdf(file="parkruncontour.pdf")
parkrun_contour()
dev.off()
```

@@

# Park run, version 2.

```{r}

rank_auck    <- c(177, 222, 206, 142, 118, 224, 128, 115, 183)
runners_auck <- c(259, 305, 297, 241, 179, 338, 203, 245, 254)

rank_stir <-    c(145, 157, 156, 191, 172, 174, 173, 165, 172, 199, 181, 229)
runners_stir <- c(193, 240, 197, 257, 227, 234, 238, 238, 196, 242, 242, 318)
```

```{r}
support_vec <- function(a,ranks,runners){
    out <- 0
    for(i in seq_along(ranks)){
        out <- out + f_single(a, ranks[i],runners[i],log=TRUE)
    }
    return(out)
}

support_ij <- function(v){
    auck <- v[1]
    stir <- v[2]
    return(
        support_vec(auck, rank_auck, runners_auck) +
        support_vec(stir, rank_stir, runners_stir)
    )
}

x <- seq(from=0.1,to=0.8,by=0.01)
jj <- as.matrix(expand.grid(x,x))
L <- apply(jj,1,support_ij)
L <- L - max(L,na.rm=TRUE)
L <- matrix(L,length(x),length(x))
```

```{r}
parkrun_contour <- function(...){
    par(pty='s')
    contour(x,x,L,asp=1,xlim=range(x),ylim=range(x),levels=-(0:5),xlab="Auckland",ylab="Stirling")
    abline(0,1)
}
parkrun_contour()
```

```{r}
M <- matrix(c(
    sum(rank_auck-1),sum(runners_auck-rank_auck),
    sum(rank_stir-1),sum(runners_stir-rank_stir)),2,2,byrow=TRUE)
dimnames(M) <- list(venue=c("Auckland","Stirling"),better=c(T,F))
M
fisher.test(M)
```
